---
output:
  pdf_document: default
  html_document: default
---
Prosper Loan Analysis by Gregory Winkler
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Packages required to run the code

library(ggplot2)
library(tidyverse)
library(lubridate)
library(GGally)
library(stringr)
library(scales)
library(maps)
library(maptools)
library(mapdata)
library(ggthemes)
library(tibble)
library(viridis)
library(readxl)



```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}

# Read in csv file containing the data
data <- read_csv("~/Documents/Udacity/eda_r/project/prosperLoanData.csv")
loanData <- data

# Read in FHLB Boston Rates
historicalRates <- read_tsv("~/Documents/Udacity/eda_r/project/historical_rates.tsv",
                            skip = 8, col_names = FALSE)
historicalRates <- subset(historicalRates, select = -c(X4))
colnames(historicalRates) <- c("Date", "Amort3y3y", "Amort5y5y")
historicalRates$Date <- as.POSIXct(strptime(historicalRates$Date, 
                                            "%m/%d/%Y", tz = 'GMT'))

# Read in Bank of America ML OAS Spread
BofAHighYieldOAS <- read_csv("~/Documents/Udacity/eda_r/project/BAMLH0A1HYBB.csv")
BofAHighYieldOAS$DATE <- as.POSIXct(BofAHighYieldOAS$DATE, 
                                                "%m/%d/%Y", tz = 'GMT')

stateAbbr <- read_excel("~/Documents/Udacity/eda_r/project/state_abbr.xlsx")

```

This report explores approximately 114,000 Propser Loan Data for loans made from
November 2005 to March 2014.  Prosper is a marketplace lending platform that 
allows for peer to peer loans.

# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}

# Adjust some of the variables to be easier to work with
loanData$IsBorrowerHomeowner <- factor(loanData$IsBorrowerHomeowner)
loanData$Occupation <- factor(loanData$Occupation)
loanData$EmploymentStatus <- factor(loanData$EmploymentStatus)
loanData$LoanStatus <- factor(loanData$LoanStatus)
loanData$CurrentlyInGroup <- factor(loanData$CurrentlyInGroup)
loanData$IncomeVerifiable <- factor(loanData$IncomeVerifiable)
loanData$Term <- factor(loanData$Term)
loanData$IncomeRange <- factor(loanData$IncomeRange,
                               levels = c("Not displayed","Not employed",
                                         "$0","$1-24,999","$25,000-49,999",
                                         "$50,000-74,999","$75,000-99,999",
                                         "$100,000+"),
                               ordered = TRUE)
loanData$LoanOriginationDate.bucket <- cut(year(loanData$LoanOriginationDate), 
                                           c(2005, 2006, 2007, 2008, 2009, 2010, 
                                             2011, 2012, 2013, 2014, 2015))

loanData$BankcardUtilization.bucket <- cut(loanData$BankcardUtilization, 
                                           c(0, 0.25, 0.5, 0.75, 1.0, 5.0))

# Display structure of the dataset
str(loanData, give.attr=FALSE, vec.len=3)
summary(loanData)

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Create buckets for grouping credit scores
CreditBuckets <- c("(400,600]", "(600,650]", "(650,675]", "(675,700]", 
                   "(700,725]", "(725,775]", "(775,900]")

# Create a new variable in the main data set that will coincide with the buckets
loanData$CreditScoreRanges <- cut(loanData$CreditScoreRangeLower, 
                                  c(400, 600, 650, 675, 700, 725, 775, 900))

```


```{r echo=FALSE, message=FALSE, warning=FALSE}

# Plot a histogram of the Lower Credit Score for each borrower. Restrict the 
# data to only include credit scores that exist and are over 400. 
ggplot(subset(loanData, 
              !is.na(CreditScoreRangeLower) & 
                CreditScoreRangeLower > 400), 
       aes(x = CreditScoreRangeLower)) + 
  geom_histogram(binwidth = 10)

# Display the mean and range of the credit score data
summary(loanData$CreditScoreRangeLower)

```

The credit scores have a median of 680.

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Plot a histogram of the Borrower Rate
ggplot(loanData, aes(x = BorrowerRate)) +
  geom_histogram(binwidth = 0.005)

# Display the mean and range of the borrower rates
summary(loanData$BorrowerRate)

```

The borrower's interest rates have a normal distribution with the exception of a 
significant number of loans around 0.33 (33%). 

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Plot a bar chart of the Income Range buckets
ggplot(loanData, aes(x = IncomeRange)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 30))

```

The largest group of borrowers have incomes in the $25,000 to $74,999 Income 
Ranges.

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Plot a histogram of the Debt-to-Income Ratios
ggplot(loanData, aes(x = DebtToIncomeRatio)) +
  geom_histogram(binwidth = 0.01) 

# Display the mean and range of the Debt-to-Income Ratios
summary(loanData$DebtToIncomeRatio)

```

The Debt-to-Income Ratio for most borrowers is reasonable with a mean around 0.27.
However, there is a sizable number of borrowers over 10.  Such a high number 
would be difficult to ever pay back. Taking a quick look at the IncomeRanges and 
whether or not that income is verifiable shows that these borrowers are 
predominantly low income and possibly may not be disclosing their true income.  

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Display the Income Range and Income Verification status of records with a 
# debt-to-income greater than 10
subset(loanData, DebtToIncomeRatio > 10) %>% 
  group_by(IncomeRange, IncomeVerifiable) %>% 
  summarise(n=n())

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Plot a histogram of the Debt-to-Income Ratios, excluding the highest 1% values
ggplot(loanData, aes(DebtToIncomeRatio)) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(limits = c(0, quantile(loanData$DebtToIncomeRatio, 0.99, 
                                            na.rm = TRUE)))

```

Excluding the highest 1% of debt-to-income ratios, we see that the majority
of borrowers are centered around 0.2.

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Plot a histogram of the Original Loan Amounts
ggplot(loanData, aes(x = LoanOriginalAmount)) + 
  geom_histogram(binwidth = 1000)

# Display the mean and range of the Original Loan Amounts
summary(loanData$LoanOriginalAmount)

```

Median loan amount is $6,500, but range from as low as $1,000 up to $35,000. 
Amounts are most frequent in $5,000 intervals.


```{r echo=FALSE, message=FALSE, warning=FALSE}

# Plot a histogram of the Monthly Loan Payments
ggplot(loanData, aes(x = MonthlyLoanPayment)) +
  geom_histogram(binwidth = 10) 

# Display the mean and range of the Monthly Loan Payments
summary(loanData$MonthlyLoanPayment)

```

Monthly loan payments are for the most part below $500. 

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Plot a histogram of the ratio of monthly loan payments to monthly stated income.
# Exclude the highest 5% of values
ggplot(loanData, aes(x = loanData$MonthlyLoanPayment / 
                       loanData$StatedMonthlyIncome)) +
  geom_histogram(binwidth = 0.0025) +
  scale_x_continuous(limits = 
                       c(0, quantile(loanData$MonthlyLoanPayment / 
                                       loanData$StatedMonthlyIncome, 0.95, 
                                     na.rm = TRUE))) +
  xlab("Monthly Loan Payment / Stated Monthly Income")

```

Dividing the monthly loan payment by the borrower's stated monthly income, 
we see that most borrowers have a monthly payment that is less than 5% of their 
income.  However there is a significant right skew to this graph. I created this
ratio because it stands to reason that borrowers paying a high amount of their 
monthly income will struggle to repay their loan.

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Plot a histogram of the Bankcard Utilization rates
ggplot(loanData, aes(x = BankcardUtilization)) +
  geom_histogram(binwidth = 0.02) +
  scale_x_continuous(limits = c(-0.1, 2))

summary(loanData$BankcardUtilization)

```

Bankcard Utilization shows a large number of borrowers that have either used 
none of their available credit or all of it.  

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Plot a histogram of the Delinquencies in the last 7 years
ggplot(loanData, aes(x = DelinquenciesLast7Years)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(limits = c(0,50)) +
  scale_y_continuous(limits = c(0,4000))

# Display a count of records with and without delinquencies
count(loanData, loanData$DelinquenciesLast7Years > 0)

```

Of the more than 36,000 borrowers that had delinquent accounts within the last 
7 years on their credit report, the majority had less than 5.  

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Plot a histogram of the Inquiries in the Last 6 Months
ggplot(loanData, aes(x = InquiriesLast6Months)) +
  geom_histogram(binwidth = 1) + 
  scale_x_continuous(limits = c(-0.5, 20), breaks = seq(0,20,1))

# Display a count of records with and without Inquiries in the Last 6 Months
count(loanData, loanData$InquiriesLast6Months > 0)

```

Many of the borrowers did not have any credit inquiries in the 6 months leading
up to their loans. 

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Plot a histogram of the Loan Origination Dates
ggplot(loanData, aes(x = LoanOriginationDate)) +
  geom_histogram(bins = 180) 

```

Most of the loans were made in 2013 and 2014, the last 2 years of the data. Note
the gap with no loans in late 2008 and first half of 2009. 

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Plot a bar chart of the Term to maturity for each loan
ggplot(loanData, aes(x = Term)) +
  geom_bar()

```

The majority of of loans are short term at 36 months to maturity.  

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Create the text list of Listing Categories
categories <- c('Not Available', 'Debt Consolidation', 'Home Improvement', 
                'Business', 'Personal Loan', 'Student Use', 'Auto', 'Other', 
                'Baby&Adoption', 'Boat', 'Cosmetic Procedure', 'Engagement Ring', 
                'Green Loans', 'Household Expenses', 'Large Purchases', 
                'Medical/Dental', 'Motorcycle', 'RV', 'Taxes', 'Vacation', 
                'Wedding Loans')

# Map the ListingCategory (numeric) column to create a new variable for the 
# text version
loanData$ListingCategory <- categories[loanData$`ListingCategory (numeric)` + 1]
loanData$ListingCategory <- factor(loanData$ListingCategory)

# Plot a bar chart of the text Listing Category
ggplot(loanData, aes(x = ListingCategory)) + 
  geom_bar() +
  theme(axis.text.x = element_text(angle=30))

```

Debt Consolidation appears as the most frequent reason for borrowing.

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Plot a bar chart for the Loan Status  
ggplot(loanData, aes(x = LoanStatus)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle=30))

```

The high charge-offs and defaulted loans in this dataset are worth taking a closer
look at to see if there are any possible causes.  Separately, the high amount of 
completed loans is expected given that the majority of loans are 36 months and 
the dataset spans just shy of 10 years.

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Create a subset of the top 15 states represented in th data
topStates <- subset(loanData, !is.na(BorrowerState)) %>% 
  group_by(BorrowerState) %>% 
  summarise(n=n()) %>% 
  arrange(desc(n)) %>% 
  top_n(15)
  
# Plot a column chart of the top 15 states
ggplot(topStates, aes(x = BorrowerState, y = n)) +
  geom_col() + 
  ylab("Count")

```

California leads the way with the most borrowers, with some of the other more 
populated states trailing behind.

# Univariate Analysis

### Observations
There are 113,937 loan records in the dataset with 81 variables about the borrower
and their loan. Some of the key variables are the borrower rate, credit score, 
debt-to-income ratio, bankcard utilization rate.

* Most of the borrowers have a credit score of approximately 680. 
* Median borrower rate is 0.184, but exhibits a right skew with a spike around 0.33.
* Median debt-to-income ratio is 0.22.
* Median bankcard utilization is 0.60 with most borrowers at 0 or 1.

### Further Exploration
I am interested in taking a closer look at the borrower's credit history to see 
what relationships there may be with a borrower's interest rate. Based on some 
of the plots we saw earlier, I would expect there to be a few variables that 
explain some of the rate differentials.  Some variables that might contribute are
the number of delinquencies, credit inquiries, debt-to-income ratio, and their 
bankcard utilization rate.  Also, some of the qualitative variables will be 
interesting to look at to see if they contributed to any disparity. For example, 
does the borrower's geography or purpose of loan have a noticeable impact.  

Also, this dataset provides an interesting time frame in that it includes loans
made at the peak of the economic cycle (2006-2007) through one of the worst 
financial collapses in recent history. I would expect defaults to be high in 2009 
and 2010 as many people became unemployed.  

The time period of this data should be an interesting factor as the financial 
collapse in 2008-2009 resulted in many people losing their jobs and therefore 
income.  I would expect defaults to be high in 2008 through 2010 as many people 
became unemployed.

### Data Adjustments
I made some adjustment to the dataset. Sepecifically, I created a new variable 
for the Listing Category to show what it is rather than a vague numerical value.
Also, I created Credit Score Buckets to group credit scores for my analysis.

# Bivariate Plots Section

A quick look at a few of the variable relationships shows some meaningful 
correlations with the borrower's credit score.  Also, the credit score appears 
to have a fairly strong negative correlation with the borrower's rate as would 
be expected.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=9, fig.height=9, Bivariate_Plots}

# Create a new dataframe with key variables to explore further
apr_factors_subset <- loanData %>% 
  select(CreditScoreRangeLower, BorrowerRate, IncomeRange,
         DebtToIncomeRatio, BankcardUtilization, CurrentDelinquencies, 
         InquiriesLast6Months)

# Create a matrix of graphs based on a few key variables
set.seed(1000)
ggpairs(apr_factors_subset[sample.int(nrow(apr_factors_subset), 5000),])

```

I quickly notice a couple of meaningful correlations for variables relative to 
the borrower's credit score.

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Plot the Bankcard Utilization rate against the Credit Scores
ggplot(subset(loanData, !is.na(CreditScoreRanges) & BankcardUtilization < 1.5), 
       aes(x = CreditScoreRanges, y = BankcardUtilization)) +
  geom_boxplot() +
  ggtitle("Bankcard Utilization by Credit Score Ranges") +
  ylab("Bankcard Utilization") +
  xlab("Credit Score Range") 

```

I took a subset of the data here to only include borrowers with a bankcard 
utilization below 1.5.  We can see the correlation between the two variables 
well here as the range and median of bankcard utilization rises as credit scores 
decline.

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Plot a boxplot of bankcard utilization and income ranges
ggplot(subset(loanData, BankcardUtilization < 1.5), 
       aes(x = IncomeRange, y = BankcardUtilization)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 30))

```

Bankcard Utilization rises slightly with income.

```{r echo=FALSE, message=FALSE, warning=FALSE}

#Plot a bar chart for current delinquencies colored by Credit Score Range
ggplot(subset(loanData, InquiriesLast6Months > 0 & !is.na(CreditScoreRanges)), 
       aes(x = CreditScoreRanges, y = InquiriesLast6Months)) +
  geom_boxplot() +
  scale_y_log10() +
  ggtitle("Credit Inquiries in the Last 6 Months - Log Scale") +
  ylab("Number of Loans Log Scale")

```

Inquiries in the last 6 months do not appear as meaningful as I was expecting. 
Typically 1 inquiry on a credit report will not cause a drop in a credit score. 
However, there does appear to be some significance to being above or below
a credit score of 700. 

```{r echo=FALSE, message=FALSE, warning=FALSE}

#Plot a bar chart for current delinquencies colored by Credit Score Range
ggplot(subset(loanData, CurrentDelinquencies > 0 & !is.na(CreditScoreRanges)), 
       aes(x = CreditScoreRanges, y = CurrentDelinquencies)) +
  geom_boxplot() +
  scale_y_log10() +
  ggtitle("Current Delinquent Accounts in Credit Report") +
  ylab("Number of Loans") 

```

Higher current delinquencies tend to be associated with lower credit scores. The
outliers for delinquencies begin to trend higher starting with borrowers below 
700 credit scores.

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Plot a dot plot of DI-to-Income Ratio to Credit Scores
ggplot(subset(loanData, DebtToIncomeRatio < 1.5 & CreditScoreRangeLower > 400), 
       aes(x = CreditScoreRanges, y = DebtToIncomeRatio)) +
  geom_boxplot() +
  geom_smooth(method = 'lm', color = 'red') +
  ggtitle("Debt-to-Income Ratio by Credit Score")

```

Debt-to-income shows normal distribution across credit scores and appears to 
have no correlation.  

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Plot a bar chart of Loan Status with color for Credit Score buckets
ggplot(subset(loanData, str_detect(LoanStatus, "Past Due")), 
       aes(x = LoanStatus, color = CreditScoreRanges)) +
  geom_bar(aes(fill=CreditScoreRanges), position = 'dodge') +
  ggtitle("Loans Past Due by Credit Score Range") +
  theme(axis.text.x = element_text(angle = 15)) +
  ylab("Number of Loans") 

```

Lower credit quality borrowers appear more likely to fall behind on 
their payments.

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Plot a violin of Defaulted Loans by Origination date and bucket by
# Credit Score Range
ggplot(subset(loanData, !is.na(CreditScoreRanges) & 
                str_detect(LoanStatus, "Defaulted") & 
                DebtToIncomeRatio < 1.1), 
       aes(x = LoanOriginationDate, y = CreditScoreRangeLower)) +
  geom_violin(aes(group = LoanOriginationDate.bucket), scale = "count") +
  ggtitle("Defaulted Loans by Origination Date and Credit Score")

```

Loan defaults were much greater in the time period leading up to the financial 
crisis then afterwards.  The lower credit score borrowers clearly were more 
likely to default prior to the crisis and again around 2011 and 2012.

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Plot the Borrower Rate by Credit Score
ggplot(subset(loanData, CreditScoreRangeLower > 400), 
       aes(x = CreditScoreRangeLower, y = BorrowerRate)) +
  geom_jitter(alpha = 0.05, size = 0.2) +
  geom_smooth(method = 'lm', color = 'red') +
  scale_y_sqrt() +
  ggtitle("Borrower Rate by Credit Score") +
  ylab("sqrt(Borrower Rate)") +
  xlab("Credit Score")

```

Credit scores show some correlation with the borrower's interest rate with an 
R^2^ of -0.46.

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Plot a boxplot of Listing Category and Borrower Rate
ggplot(loanData, aes(x = ListingCategory, y = BorrowerRate)) +
  geom_boxplot() + 
  xlab("Listing Category") +
  ylab("Borrowing Rate") +
  ggtitle("Borrowing Rates by Loan Purpose") +
  theme(axis.text.x = element_text(angle=30))

```

Loan purposes listed as Not Available have the greatest outliers. Cosmetic 
procedures and Household expenses have the highest rates on average.

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Plot a bokxplot of Borrower State and Borrower Rate
ggplot(subset(loanData, str_detect(BorrowerState, c(topStates$BorrowerState))), 
       aes(x = BorrowerState, y = BorrowerRate)) +
  geom_boxplot() + 
  xlab("Borrower State") +
  ylab("Borrowing Rate") +
  ggtitle("Borrowing Rates by State")

```

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Create a new table for defaults by state and plot a US map of the data
states_map <- map_data("state")
stateData <- subset(loanData, !is.na(BorrowerState) & BorrowerState != "DC") %>% 
  group_by(BorrowerState) %>% 
  summarise(defaultPerc = sum(str_detect(LoanStatus, "Defaulted"))/n())

stateData <- left_join(stateData, stateAbbr, c("BorrowerState" = "Abbreviation:"))
colnames(stateData) <- c("abbreviation","defaultPerc","state")
stateData$state <- tolower(stateData$state)

stateData$defaultPerc.bucket <- cut(stateData$defaultPerc,
                            c(0.000,0.025,0.05,0.075,0.1,1))


ggplot(stateData, aes(map_id = state)) + 
  geom_map(aes(fill = defaultPerc.bucket), map = states_map, color ="black") +
  expand_limits(x = states_map$long, y = states_map$lat) +
  ggtitle("% of Loans in Default by State")


```

As a percentage of the total loans in a state, the northeast region looks to be 
the least likely area to default, whereas the northwest and midwest show the
greatest potential for default.

# Bivariate Analysis

### Observations
It was interesting to note some of the variables that had a correlation with the
credit score. As expected, the bankcard utilization, current delinquencies and 
number of inquiries in the last 6 months all exhibited some correlation. However,
I was a little surprised that the debt-to-income ratio showed no correlation. 
While income may not factor into the credit score, I would have expected an 
indirect relationship here where borrowers that have high debt-to-income are
more likely to fall behind on payments and/or utilize a large portion of their 
available credit.  

Near the beginning of the analysis it was noted that a spike in borrowing rates
was observed around 33%.  This appeared with some surprising results when graphed 
against credit scores.  The data showed a negative correlation between credit 
scores and borrower rates, but for the subset of data with rates around 33%
showed a high concentration corresponding to borrowers with credit scores
above what might be expected.  Based on the graph of the relationship, I would 
not have expected many loan rates to coincide with credit scores over 700.

The number of credit inquiries and current delinquencies both showed that a 
higher occurrence in either was more likely to be associated with borrowers 
having lower credit scores.


Defaults were more frequent among low credit score borrowers which was to be 
expected. However, the more interesting aspect of the graph was the impact that
exogenious factors have on the repayment of a loan. While a credit score can 
quantify a borrower's ability to repay their loan based on historical information,
it is unable to anticipate future variants that can lead to unexpected results. 
This is evident in the high frequency of defaults preceeding the 2008-2009 
financial crisis when many people became unemployed. 

The rise in borrower rates in 2011 through 2012 was interesting to note beause
the FHLB Boston 3 year fully amortizing rate, a benchmark rate used by commercial 
banks, stayed at very low levels during this period.  Since credit scores are
representative of certain characteristic traits, it means that either lenders
were demanding a higher risk premium or borrowers were taking loans out for 
longer terms.


The strongest relationship in the data I explored was the Credit Score and the 
Borrower's Rate with a R^2^ of -0.46. The Bankcard Utilization rate and the Credit 
Scores was not significantly different though with a R^2^ of -0.40. Also, Current 
Delinquencies and Credit Inquiries showed some correlation with Credit Scores.

# Multivariate Plots Section

```{r echo=FALSE, warning=FALSE, message=FALSE, Multivariate_Plots}

#Plot a dotplot of Credit Scores and Borrower rate with colro for Bankcard Utilization 
ggplot(data = subset(loanData, CreditScoreRangeLower > 400 & 
                       BankcardUtilization < 5.0 & 
                       !is.na(BankcardUtilization)), 
       aes(x = factor(CreditScoreRangeLower), 
           y = BorrowerRate, 
           fill = BankcardUtilization.bucket)) + 
  geom_boxplot() +
  theme(legend.position = 'bottom')

```

In general, a higher bankcard utilization rate corresponds with lower credit 
scores and higher interest rates.

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Plot a boxplot of Credit scores and borrower rate for loans past due
ggplot(subset(loanData, !is.na(BankcardUtilization.bucket) & CreditScoreRangeLower > 400 & 
                BankcardUtilization < 5.0 & 
                str_detect(LoanStatus, 'Past Due')==TRUE), 
       aes(x = CreditScoreRanges, y = BorrowerRate, 
           fill = BankcardUtilization.bucket)) +
  geom_boxplot() +
  ggtitle("Loans Past Due")

```

Loans that are past due tend to be higher rates and borrowers that have low
credit scores and high bankcardutilization.

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Plot a histogram of loans with rates between 31% and 34% having a credit score
# over 700
ggplot(subset(loanData, CreditScoreRangeLower > 700 & BorrowerRate > 0.31 &
                BorrowerRate < 0.34), 
       aes(x = LoanOriginationDate, color = CreditScoreRanges)) +
  geom_histogram(bins = 60, aes(fill=CreditScoreRanges)) +
  ylab("Number of Loans") +
  ggtitle("Loans with Borrower Rates between 31% and 34%")

```

Graphing only the loans with interest rates between 31% and 34% having a credit
score greater than 700, we see that these loans were almost entirely originated
in 2011 and 2012.  

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Group and get the median rate by Credit Score and Loan Origination Date
RatesByDayAndCredit <- loanData %>% 
  group_by(LoanOriginationDate, CreditScoreRanges) %>% 
  filter(!is.na(CreditScoreRanges)) %>% 
  summarise(medianRate = median(BorrowerRate),
            n = n())

colnames(RatesByDayAndCredit)[1] <- c("Date")

# Add external interest rate data to the dataframe
RatesByDayAndCredit <- left_join(RatesByDayAndCredit, historicalRates)
RatesByDayAndCredit <- left_join(RatesByDayAndCredit, BofAHighYieldOAS,
                                 c("Date" = "DATE"))
colnames(RatesByDayAndCredit)[7] <- c("BofA_HY_OAS")

# Plot a smoothed line graph of the median borrowing rate by credit score range
ggplot(RatesByDayAndCredit, 
       aes(x = Date, y = medianRate, color = CreditScoreRanges)) +
  geom_smooth() +
  geom_smooth(aes(x = Date, y = Amort3y3y / 100, color = 'FHLB 3y Rate')) +
  geom_smooth(aes(x = Date, y = (BofA_HY_OAS + Amort3y3y) / 100, 
                  color = '3 Yr Rate + BofA HY BB OAS')) 

```

Graphing the median borrowing rates over time, we see that interest rates started
rising in 2010 and remained elevated through most of 2012. Meanwhile the FHLB
Boston 3 year amortizing rate, a lending benchmark rate, remained at very low 
levels.

There is a slight increase in the credit risk premium in 2011 that may explain 
some of the rise in borrower rates observed. 

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Plot a smoothed line of loans per day by credit score
ggplot(RatesByDayAndCredit, aes(x = Date, y = n)) +
  geom_smooth(aes(color = CreditScoreRanges)) + 
  ggtitle("Number of Loans per day by Credit Score") +
  ylab("Number of Loans") +
  xlab("Loan Origination Date")

```

Separately we see a significant increase in loans to borrowers with 675 to 700
credit scores starting in 2011 and increasing at a significantly faster pace in
2013.

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Plot a violin of loan origination date and borrower rate, with separate graphs 
# for each term
ggplot(subset(loanData, !is.na(CreditScoreRanges)), 
              aes(x = LoanOriginationDate, y = BorrowerRate)) +
  geom_violin() +
  facet_wrap(~Term)

```

It appears that 12 and 60 month loans are new to Prosper starting around 2011. 

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Plot a dotplot as the previous one but with only credit scores between 
# 700 and 775
ggplot(subset(loanData, CreditScoreRangeLower > 700 & 
                CreditScoreRangeLower <= 775 & Term != 12), 
       aes(x = LoanOriginationDate, y = BorrowerRate, 
           color = CreditScoreRanges)) +
  geom_jitter(size = 0.5) +
  facet_wrap(~Term)

```

There has been an increase in 60 month term loans for borrowers with credit 
scores ranging from 700 to 775.  However, the concentration around the 33% 
interest rate looks to be 36 month loans. 

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Plot a histogram of the loan originations with their term
ggplot(subset(loanData, LoanOriginationDate > "2011-01-01"), 
              aes(x = LoanOriginationDate, color= Term)) +
  geom_histogram(aes(fill = Term), bins = 60)

```

The 60 month loans do not make a noticable increase until 2012.

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Plot a density graph of bankcard utilization with credit scores and income range
ggplot(subset(loanData, CreditScoreRangeLower > 700 & 
                CreditScoreRangeLower <= 775 &
                BorrowerRate > 0.31 & BorrowerRate < 0.34), 
       aes(x = BankcardUtilization, color = CreditScoreRanges)) +
  geom_density() +
  facet_wrap(~IncomeRange) +
  ggtitle("Bankcard Utilization of Borrowers with Rates between 31% and 34%")


```

For the selected range of credit scores, it does appear that bankcard utilization
for higher income borrowers may be a contributing factor for the higher rates.  

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Plot a density graph of debt-to-income with credit scores and income range
ggplot(subset(loanData, CreditScoreRangeLower > 700 & 
                CreditScoreRangeLower <= 775 & BorrowerRate > 0.31 & 
                BorrowerRate < 0.34 & DebtToIncomeRatio < 1.1), 
       aes(x = DebtToIncomeRatio, color = CreditScoreRanges)) +
  geom_density() +
  facet_wrap(~IncomeRange) +
  ggtitle("Debt-to-Income of Borrowers with Rates between 31% and 34%")


```

Whereas higher debt-to-income may be more likely a contributing factor for lower 
income borrowers.

# Multivariate Analysis

### Observations

The median interest rates of the credit score ranges over the relevent period 
was insightful to provide a possible reason for why the data did not necessarily
fit as well with other variables as I might have expected. Considering many of 
these borrowers are high risk, I would not be surprised if the increase in 
overall borrower rates during 2011 was from lenders being more conservative and
expecting a higher return.  

While bankcard utilization varied greatly, in general it was noticeably higher 
for loans that had a low credit score and high interest rate.  Also, it looks to 
be a contributing factor to loans that are currently past due. This was inline 
with my expectations as these borrowers were most likely already struggling with 
repaying their debt.

As interest rate declined in 2013 and 2014, borrowers took advantage of the 60
month term option in increasing numbers.  

### Interesting Interactions between Features

It was interesting to see that for the interest rates between 31% and 34% of 
higher credit score borrowers, bankcard utilization had a bigger factor for 
higher income borrowers.  This was in contrast to debt-to-income that appeared
to be more influential for lower income borrowers.  

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}

# Plot a violin of Loan origination date and Credit scores
ggplot(subset(loanData, !is.na(CreditScoreRanges) & 
                str_detect(LoanStatus, "Defaulted") & 
                DebtToIncomeRatio < 1.1), 
       aes(x = LoanOriginationDate, y = CreditScoreRangeLower, 
           color = LoanOriginationDate.bucket)) +
  geom_violin(aes(fill = LoanOriginationDate.bucket), scale = "count") +
  ggtitle("Defaulted Loans by Origination Date and Credit Score") +
  ylab("Credit Score") +
  xlab("Loan Origination Date")

```

### Description One

The difference in number of defaulted loans before and after the financial crisis
was very interesting.  This shows how our models and research can only provide
insight on what we can expect based on historical examples. Also, we can see that 
these borrowers are of lower quality than the overall dataset with a median 
credit score of 640 and borrowing rate of 22.7%.  

**Credit Scores**

| Min.   | 1st Qu. | Median | Mean   | 3rd Qu. | Max.   |
|--------|---------|--------|--------|---------|--------|
| 420.0  | 560.0   | 640.0  | 625.3  | 680.0   | 860.0  |
	
**Borrower Rates**

| Min.   | 1st Qu. | Median | Mean   | 3rd Qu. | Max.   |
|--------|---------|--------|--------|---------|--------|
| 0.00%  | 16.50%  | 22.70% | 22.15% | 28.50%  | 36.00% |


### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}

# Plot a boxplot of Credit scores and Borrower rate
ggplot(data = subset(loanData, CreditScoreRangeLower > 400 & 
                       BankcardUtilization < 5.0 & 
                       !is.na(BankcardUtilization)), 
       aes(x = factor(CreditScoreRangeLower), 
           y = BorrowerRate, 
           fill = BankcardUtilization.bucket)) + 
  geom_boxplot( ) +
  ggtitle("Bankcard Utilization by Credit Score and Borrower Rate") +
  xlab("Credit Score") +
  ylab("Borrower Rate") +
  theme(legend.position = 'bottom') +
  guides(fill=guide_legend(title="Bankcard Utilization Buckets"))

```

### Description Two

The relationship between credit scores and borrower rates shows that there is a 
negative correlation (R^2^ of -0.46)  This makes sense as we would expect that 
the reason for the borrower's low credit score is that they would be more likely 
to not pay back their loan in full.

| Credit Score Bucket | Median Bankcard Utilization |
| ------------------- | --------------------------- |
| (400,600]	          | 0.86   	                    |	
| (600,650]	          | 0.78   	                    |	
| (650,675]	          | 0.73   	                    |	
| (675,700]	          | 0.65   	                    |	
| (700,725]	          | 0.52   	                    |	
| (725,775]	          | 0.38   	                    |	
| (775,900]	          | 0.18   	                    |	

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}

# Plot a smoothed line for each credit score bucket of median borrower rates
ggplot(RatesByDayAndCredit, 
       aes(x = Date, y = medianRate, color = CreditScoreRanges)) +
  geom_smooth() +
  geom_smooth(aes(x = Date, y = Amort3y3y / 100, color = 'FHLB 3y Rate')) +
  geom_smooth(aes(x = Date, y = (BofA_HY_OAS + Amort3y3y) / 100, 
                  color = '3 Yr Rate + BofA HY BB OAS')) +
  scale_y_continuous(labels = percent) +
  ylab("Median Borrower Rate") +
  ggtitle("Median Rate by Credit Score over the Sample Period")

```

### Description Three

The median interest rates by credit score over time provide a couple of useful
pieces of information. The show that lower credit borrowers are likely to have
higher interest rates at any given point in time. Additionally, they show that 
during time periods of higher risk, these low credit borrowers may see larger 
differentials in pricing from other borrowers.

| Credit Score Bucket | Median Rate |
| ------------------- | ----------- |
| (400,600]	          | 25.00%	    |		
| (600,650]	          | 23.20%			|
| (650,675]	          | 20.99%			|
| (675,700]	          | 18.25%			|
| (700,725]	          | 15.20%			|
| (725,775]	          | 13.50%			|
| (775,900]	          | 9.99% 	    |

------

# Reflection

The Prosper dataset was interesting to explore as there was information about 
the borrower's historical experience with debt and also their current loan and
its repayment performance.  It was relatively easy to work with the data as 
most of it was already in a workable format.  However, there were a few 
variables that I found helpful to factor first to make them easier to work with.

It was interesting to bring in external interest rate data into the dataset. The
Prosper data provided a lot of interesting information but there were a couple 
of times that I thought a stronger relationship should have existed but did not. 
Understanding the general interest rate market conditions helped to guide further
exploration.  This was most helpful with exploring the large spike in loan rates
around 33%.  

In the future, I think the analysis could be expanded to try to predict interest
rates on new loans.  I think there is enough information from the propser data
that when combined with general market yield curves, it could come close to 
providing some prediction on where new loans would be given a set of borrower
characteristics.  However, Prosper's business model likely makes this a little 
more difficult in that at any point in time there could be an imbalance of 
borrowers and lenders since the business is not universally known as a typical
bank would be.  


### Additional data sources

1. FHLB Boston Rates: http://www.fhlbboston.com/rates/historicalrates/index.jsp
2. Bank of America Merrill Lynch OAS: https://fred.stlouisfed.org/series/BAMLH0A1HYBB